---
title: "Ancestry in America"
subtitle: "Three views of the same ACS ancestry data"
"author": "Diya Sharma"
format:
  html:
    embed-resources: true
    toc: true
    fig-width: 13
    fig-height: 8.5
    fig-format: png
    fig-dpi: 300
execute:
  echo: false
  warning: false
  message: false
  cache: true
---

```{r}
#| label: libraries

library(tidyverse)
library(tidycensus)
library(tigris)
library(sf)
library(leaflet)
library(htmltools)
library(glue)
library(scales)

options(tigris_use_cache = TRUE)
```

```{r}
#| label: fetch-data

yr <- 2023  # most recent ACS 5-year vintage

ancestry_raw <- get_acs(
  geography = "county",
  table     = "B04006",
  year      = yr,
  survey    = "acs5"
)

vars <- load_variables(yr, "acs5", cache = TRUE) %>%
  filter(str_starts(name, "B04006"))
```

```{r}
#| label: fetch-race

race_raw <- get_acs(
  geography = "county",
  table     = "B02001",
  year      = yr,
  survey    = "acs5"
)

race_pct <- race_raw %>%
  filter(variable %in% c("B02001_001", "B02001_002")) %>%
  select(GEOID, variable, estimate) %>%
  pivot_wider(names_from = variable, values_from = estimate) %>%
  mutate(pct_white = B02001_002 / B02001_001) %>%
  select(GEOID, pct_white)
```

```{r}
#| label: clean-labels

ancestry <- ancestry_raw %>%
  left_join(vars %>% select(variable = name, label), by = "variable") %>%
  mutate(
    group_raw = label %>%
      str_remove("Estimate!!Total:!!") %>%
      str_remove("^Total:$") %>%
      str_trim()
  ) %>%
  filter(
    !variable %in% c("B04006_001"),
    !str_detect(group_raw, regex("other groups|not reported|unclassified",
                                  ignore_case = TRUE)),
    group_raw != ""
  )

total_pop <- ancestry_raw %>%
  filter(variable == "B04006_001") %>%
  select(GEOID, total = estimate)
```

```{r}
#| label: crosswalk

ancestry_groups <- tribble(
  ~pattern,                                              ~group,
  "^American$",                                          "American",
  "German",                                               "German",
  "Irish$",                                                "Irish",
  "Scotch-Irish",                                          "Scotch-Irish",
  "^Scottish",                                             "Scottish",
  "English",                                               "English",
  "^Welsh",                                                "Welsh",
  "British",                                               "British",
  "French(?! Canadian)",                                   "French",
  "French Canadian",                                       "French Canadian",
  "Italian",                                                "Italian",
  "Polish",                                                 "Polish",
  "Russian",                                                "Russian",
  "Ukrainian",                                              "Ukrainian",
  "Swedish",                                                "Swedish",
  "Norwegian",                                              "Norwegian",
  "Danish",                                                 "Danish",
  "Dutch",                                                  "Dutch",
  "Finnish",                                                "Finnish",
  "Hungarian",                                              "Hungarian",
  "Czech",                                                  "Czech",
  "Greek",                                                  "Greek",
  "Portuguese",                                             "Portuguese",
  "Arab",                                                   "Arab",
  "Armenian",                                               "Armenian",
  "Subsaharan African",                                     "Subsaharan African",
  "West Indian",                                            "West Indian",
  "Mexican",                                                "Mexican",
  "Puerto Rican",                                           "Puerto Rican",
  "Cuban",                                                  "Cuban",
  "Slavic",                                                 "Slavic (other)"
)

assign_group <- function(x) {
  result <- rep(NA_character_, length(x))
  for (i in seq_len(nrow(ancestry_groups))) {
    idx <- which(is.na(result))
    if (length(idx) == 0) break
    hits <- str_detect(x[idx], regex(ancestry_groups$pattern[i], ignore_case = TRUE))
    result[idx[hits]] <- ancestry_groups$group[i]
  }
  result
}

ancestry <- ancestry %>%
  mutate(group = assign_group(group_raw)) %>%
  filter(!is.na(group))

ancestry_grouped <- ancestry %>%
  group_by(GEOID, NAME, group) %>%
  summarise(estimate = sum(estimate, na.rm = TRUE), .groups = "drop")
```

```{r}
#| label: palette

palette <- c(
  "American"            = "#a6611a",
  "German"               = "#dfc27d",
  "Irish"                 = "#80cdc1",
  "Scotch-Irish"          = "#018571",
  "Scottish"              = "#35978f",
  "English"               = "#c7eae5",
  "Welsh"                 = "#8dd3c7",
  "British"               = "#b2dfdb",
  "French"                = "#9ecae1",
  "French Canadian"       = "#4292c6",
  "Italian"               = "#e08214",
  "Polish"                = "#8073ac",
  "Russian"               = "#542788",
  "Ukrainian"             = "#b35806",
  "Swedish"               = "#fee391",
  "Norwegian"             = "#fec44f",
  "Danish"                = "#fe9929",
  "Dutch"                 = "#fdae6b",
  "Finnish"               = "#a1d99b",
  "Hungarian"             = "#c51b7d",
  "Czech"                 = "#de77ae",
  "Greek"                 = "#7fbc41",
  "Portuguese"            = "#4d9221",
  "Arab"                  = "#276419",
  "Armenian"              = "#c2a5cf",
  "Subsaharan African"    = "#d6604d",
  "West Indian"           = "#f4a582",
  "Mexican"               = "#67001f",
  "Puerto Rican"          = "#b2182b",
  "Cuban"                 = "#d1e5f0",
  "Slavic (other)"        = "#92c5de"
)
```

```{r}
#| label: top-groups

top5 <- ancestry_grouped %>%
  left_join(total_pop, by = "GEOID") %>%
  mutate(pct = estimate / total) %>%
  group_by(GEOID) %>%
  arrange(desc(estimate), .by_group = TRUE) %>%
  mutate(rank = row_number()) %>%
  filter(rank <= 5) %>%
  ungroup()

plurality <- top5 %>%
  filter(rank == 1) %>%
  select(GEOID, NAME, top_group = group, total, top_pct = pct)
```

```{r}
#| label: tooltips

build_tooltip <- function(name, total, pct_white, groups_df) {
  rows_html <- groups_df %>%
    mutate(row = glue(
      '<div style="display:flex;align-items:center;margin:3px 0;">
         <span style="width:10px;height:10px;background:{palette[group]};
                      display:inline-block;margin-right:7px;border-radius:2px;flex-shrink:0;"></span>
         <span style="flex:1;">{group}</span>
         <span style="font-weight:600;margin-left:10px;">{percent(pct, accuracy = 1)}</span>
       </div>'
    )) %>%
    pull(row) %>%
    paste(collapse = "")

  white_line <- if (!is.na(pct_white)) {
    glue('<div style="color:#bbb;font-size:11px;margin-top:8px;">',
         'Overall, {percent(pct_white, accuracy = 1)} identified as white.</div>')
  } else ""

  glue(
    '<div style="font-family:-apple-system,sans-serif;min-width:230px;padding:2px;">
       <div style="font-weight:700;font-size:14px;margin-bottom:2px;">{name}</div>
       <div style="color:#bbb;font-size:12px;margin-bottom:6px;">{comma(total)} people reporting ancestry</div>
       <hr style="border-color:#555;margin:4px 0;">
       {rows_html}
       {white_line}
     </div>'
  )
}

tooltip_lookup <- top5 %>%
  left_join(race_pct, by = "GEOID") %>%
  group_by(GEOID, NAME, total, pct_white) %>%
  nest() %>%
  mutate(tooltip = pmap_chr(
    list(NAME, total, pct_white, data),
    build_tooltip
  )) %>%
  ungroup() %>%
  select(GEOID, tooltip)
```

```{r}
#| label: geometry

county_geo <- counties(cb = TRUE, resolution = "20m", year = yr) %>%
  select(GEOID, geometry) %>%
  st_transform(4326)   # leaflet requires WGS84

map_data <- county_geo %>%
  left_join(plurality, by = "GEOID") %>%
  left_join(tooltip_lookup, by = "GEOID") %>%
  filter(!is.na(top_group))
```

```{r}
#| label: state-geometry

state_geo <- states(cb = TRUE, resolution = "20m", year = yr) %>%
  filter(!STUSPS %in% c("PR", "VI", "GU", "AS", "MP")) %>%
  select(STUSPS, geometry) %>%
  st_transform(4326)
```

## Map 1 — The largest ancestry group in every county

Hover any county for its top 5 reported ancestry groups by share of the
ancestry-reporting population, plus the share of the county's population
that identified as white overall (ACS table `B02001`).

```{r}
#| label: fig-map-plurality
#| column: page

levels_present <- intersect(names(palette), unique(map_data$top_group))
pal <- colorFactor(palette = unname(palette[levels_present]), domain = levels_present)

leaflet(map_data, options = leafletOptions(minZoom = 3)) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  setView(lng = -96, lat = 38, zoom = 4) %>%
  addPolygons(
    fillColor   = ~pal(top_group),
    fillOpacity = 0.75,
    color       = "white",
    weight      = 0.4,
    smoothFactor = 0.3,
    highlightOptions = highlightOptions(
      weight = 2, color = "#333", fillOpacity = 0.9, bringToFront = TRUE
    ),
    label = ~lapply(tooltip, HTML),
    labelOptions = labelOptions(
      style = list(
        "background-color" = "rgba(20,20,20,0.94)",
        "color" = "white",
        "border" = "none",
        "box-shadow" = "0 2px 10px rgba(0,0,0,0.35)"
      ),
      textsize  = "13px",
      direction = "auto",
      opacity   = 1
    )
  ) %>%
  addPolygons(
    data = state_geo,
    fill = FALSE,
    color = "#222222",
    weight = 1.3,
    opacity = 0.8
  ) %>%
  addLegend(
    pal = pal, values = ~top_group,
    title = "Largest ancestry group",
    position = "bottomright"
  )
```

## Map 2 — How mixed is each county's ancestry?

The plurality map above only shows who's *biggest* — it treats a county
that's 90% German the same as one that's 25% German, 24% Irish, 23%
American, and a long tail of everything else, as long as German edges
out narrowly in both. This map scores every county on how *evenly split*
its reported ancestry is, using a Herfindahl–Hirschman-style diversity
index: for each county, square each group's share of the ancestry-
reporting population and sum those squares (this "concentration score"
is high when one group dominates), then subtract from 1. A score near 0
means one group overwhelmingly dominates; a score near 1 means ancestry
is spread thinly across many groups with no clear plurality. This is a
measure of *reported-ancestry* concentration specifically, not a general
racial or ethnic diversity index.

```{r}
#| label: diversity-index

diversity <- ancestry_grouped %>%
  left_join(total_pop, by = "GEOID") %>%
  mutate(pct = estimate / total) %>%
  group_by(GEOID) %>%
  summarise(hhi = sum(pct^2, na.rm = TRUE), .groups = "drop") %>%
  mutate(diversity_index = 1 - hhi)

diversity_map_data <- map_data %>%
  left_join(diversity, by = "GEOID") %>%
  mutate(
    tooltip2 = glue(
      '<div style="font-family:-apple-system,sans-serif;min-width:200px;">
         <div style="font-weight:700;font-size:14px;margin-bottom:4px;">{NAME}</div>
         <div>Diversity index: <strong>{round(diversity_index, 2)}</strong></div>
         <div style="color:#bbb;font-size:12px;margin-top:4px;">
           Largest group: {top_group} ({percent(top_pct, accuracy = 1)})
         </div>
       </div>'
    )
  )
```

```{r}
#| label: fig-map-diversity
#| column: page

pal_div <- colorNumeric(palette = "viridis", domain = diversity_map_data$diversity_index)

leaflet(diversity_map_data, options = leafletOptions(minZoom = 3)) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  setView(lng = -96, lat = 38, zoom = 4) %>%
  addPolygons(
    fillColor   = ~pal_div(diversity_index),
    fillOpacity = 0.8,
    color       = "white",
    weight      = 0.3,
    smoothFactor = 0.3,
    highlightOptions = highlightOptions(
      weight = 2, color = "#333", fillOpacity = 0.95, bringToFront = TRUE
    ),
    label = ~lapply(tooltip2, HTML),
    labelOptions = labelOptions(
      style = list(
        "background-color" = "rgba(20,20,20,0.94)",
        "color" = "white",
        "border" = "none",
        "box-shadow" = "0 2px 10px rgba(0,0,0,0.35)"
      ),
      textsize  = "13px",
      direction = "auto",
      opacity   = 1
    )
  ) %>%
  addPolygons(
    data = state_geo,
    fill = FALSE,
    color = "#222222",
    weight = 1.3,
    opacity = 0.8
  ) %>%
  addLegend(
    pal = pal_div, values = ~diversity_index,
    title = "Ancestry diversity index",
    position = "bottomright"
  )
```

## Map 3 — Where six individual groups are concentrated

The first two maps only ever show each county's single largest or most
evenly split group, which erases everything else happening underneath.
This view flips that around: six small maps, each shaded by *one*
group's share of the local population, so you can see where German,
Irish, Mexican, Italian, Subsaharan African, and self-identified
"American" ancestry are concentrated — independent of whether that group
happens to be the county's plurality winner.

```{r}
#| label: fig-map-facets
#| fig-width: 13
#| fig-height: 9.5
#| column: page

selected_groups <- c("American", "German", "Irish", "Mexican", "Italian", "Subsaharan African")

facet_data <- expand_grid(GEOID = unique(total_pop$GEOID), group = selected_groups) %>%
  left_join(
    ancestry_grouped %>% filter(group %in% selected_groups),
    by = c("GEOID", "group")
  ) %>%
  mutate(estimate = replace_na(estimate, 0)) %>%
  left_join(total_pop, by = "GEOID") %>%
  mutate(pct = estimate / total)

static_geo <- county_geo %>%
  shift_geometry()   # Albers projection with AK/HI inset, for a static print-style map

facet_map_data <- static_geo %>%
  left_join(facet_data, by = "GEOID") %>%
  filter(!is.na(group)) %>%
  mutate(group = factor(group, levels = selected_groups))

ggplot(facet_map_data) +
  geom_sf(aes(fill = pct), color = NA) +
  facet_wrap(~group, ncol = 3) +
  scale_fill_viridis_c(
    option = "magma", trans = "sqrt", labels = percent,
    name = "Share of county population",
    guide = guide_colorbar(barwidth = 18, barheight = 0.6)
  ) +
  coord_sf(datum = NA) +
  theme_void(base_size = 11) +
  theme(
    strip.text = element_text(face = "bold", size = 11),
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(size = 10, hjust = 0.5, color = "grey30")
  ) +
  labs(
    title = "Where Six Ancestry Groups Live in the U.S.",
    subtitle = paste0("Share of county population reporting each ancestry \u2022 ACS ", yr - 4, "\u2013", yr, " 5-year estimates"),
    caption = "Source: U.S. Census Bureau, ACS Table B04006. Color scale uses a square-root transform to make lower-share concentration visible."
  )
```

## Notes on interpretation

- **Map 1** uses the same crosswalk from ~90 detailed ACS ancestry codes
  into ~30 broad display groups discussed in Map 3; **American** ancestry
  is kept as its own category rather than folded into English/British,
  since in Appalachia and the South it reflects a distinct
  self-identification, not missing data.
- **Map 2**'s diversity index is a property of *reported ancestry
  categories*, not a general measure of racial, ethnic, or cultural
  diversity — a county could score low here (one dominant reported
  ancestry) while still being demographically diverse in other respects.
- **Map 3** uses a square-root color transform rather than a linear one,
  which compresses the visual difference between high-share counties and
  stretches out the low end — useful here because most of these
  distributions are highly skewed (a few very concentrated counties, a
  long tail of near-zero counties).
- All three maps are **county-level** (ACS 5-year, `r yr`) for
  performance across the full country. For census-tract precision within
  a single metro, scope `geography = "tract"` to one `state =` in the
  data-fetching steps.
